Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-14 _tex/index.tex
Built site for gh-pages

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might seek to answer
the question, ``To what extent do daily practice hours (\(X\)) influence
students' ability to produce high-quality written texts (\(Y\))?''. To
investigate this, the study randomly assigns students to two groups with
different levels of daily practice, low \((X=1)\) and medium \((X=2)\).
The quality of the written texts is evaluated using pairwise
comparisons, resulting in a dichotomous outcome \((Y=\{0,1\})\). The
research question is then framed as the estimand, ``\emph{On average},
which group of students produces higher-quality written texts?'' This
estimand is mathematically expressed as \(E[Y|X=1] - E[Y|X=2]\), where
\(E[\cdot]\) represents the expected value.

Researchers then proceed to identify the estimands.
\emph{Identification} is the process of accurately computing an estimand
using an estimator. An \emph{estimator} is a method or function that
transforms data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which is the process that integrates data with an
estimator \citep{Everitt_et_al_2010}. While various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure accurate estimates. For example,
Fisher's exact test \citep{Fisher_1922} is an estimator with desirable
properties for comparing \(2\times2\) tables. The test provides accurate
estimates for comparing proportions when its assumptions are satisfied.
Figure~\ref{fig-IEflow}, known as the Identification-Estimation
flowchart \citep{McElreath_2020, Neal_2020}, illustrates the
aforementioned process of transitioning from estimands to estimates.

@@

However, many studies aim to establish causal relationships rather than
merely infer associations. In the earlier example, the estimates
obtained using Fisher's exact test can be interpreted as causal because
the data were collected through a randomized experiment. In randomized
experiments, the researcher fully controls the treatment assignment
mechanism, such as assigning students to different groups. This control
allows causal estimates to be derived from associational estimates, such
as those provided by the Fisher statistic, since randomization
effectively eliminates confounding \citep{Neal_2020}. \emph{Confounding}
occurs when an external variable influences both the outcome and the
variable of interest, leading to spurious association
\citep{Everitt_et_al_2010}.

>>>>

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might examine the
question, ``To what extent do different teaching methods (\(X\))
influence students' conceptual understanding of algebra (\(Y\))?'' To
investigate this, the study randomly assigns students to two groups,
each using different algebra teaching method \((X=\{1,2\})\). Students'
conceptual understanding of algebra is evaluated through pairwise
comparisons, resulting in a dichotomous outcome \((Y=\{0,1\})\),
indicating which student possesses a higher degree of understanding
among those compared. The research question is then framed as the
estimand, ``\emph{On average}, is there a difference in conceptual
understanding of algebra between the two groups of student?'' This
estimand is mathematically expressed as \(E[Y| X=1] - E[Y| X=2]\), where
\(E[\cdot]\) denotes the expected value. An example of this approach is
seen in \citet{Jones_et_al_2019}.

Researchers then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which refer to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. Although various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure the accuracy of estimates. For
instance, the BTL model \citep{Bradley_et_al_1952, Luce_1959} is an
estimator known for its effectiveness in modeling pairwise comparisons,
providing accurate estimates for the desired estimand when its
assumptions are met.

However, many studies aim to understand the mechanisms underlying
specific data types and also seek to establish causal relationships
rather than merely infer associations. In the earlier example, the
associational estimates obtained using the BTL model can be interpreted
as causal, because the data were collected using a randomized
experiment. Randomized experiments enable the causal interpretation of
associational estimates by ensuring several key properties, including
common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most crucial property,
however, is that randomization effectively eliminates confounding.
\emph{Confounding} occurs when an external variable influences both the
outcome and the variable of interest, leading to spurious association
\citep{Everitt_et_al_2010}. Randomization mitigates this issue by
decoupling the intervention assignment mechanism, such as assigning
students to different groups, from other variables and outcomes
\citep{Morgan_et_al_2014, Neal_2020}.

Although experiments are widely regarded as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014},
researchers often encounter constraints that limit their ability to
conduct experimental studies. These constraints include ethical
concerns, such as the assignment of individuals to potentially harmful
interventions, and practical limitations, such as the infeasibility of,
for example, assigning individuals to genetic modifications or physical
impairments \citep{Neal_2020}. In these instances, causal inference
offers a valuable alternative for generating causal estimates.
\emph{Causal inference} is a framework designed to identify the causes
of phenomena and estimate their effects using data
\citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical statistical
modeling, which focuses primarily on summarizing data and inferring
associations, causal inference provides a rigorous mathematical
framework for analyzing causes and counterfactuals \citep{Pearl_2009}.
Moreover, the framework is beneficial for both observational and
experimental studies, as it provides significant theoretical insights
that enhance their design \citep{McElreath_2020}.

Counterfactuals represent the highest level of cognitive abstraction in
the ladder of causation, followed by intervention and association
\citep{Pearl_et_al_2018}, and a detailed explanation of causes,
counterfactuals, and their applications is beyond the scope of this
document. However, it is important to note that causal inference
addresses the question, ``What would have happened in the world under
different circumstances?'' by transforming unobserved \emph{potential
outcomes}, known as counterfactuals, into associational estimates.
\emph{Counterfactuals} represent scenarios \emph{contrary to fact},
where alternative \emph{potential outcomes} resulting from a cause are
neither observed nor observable \citep{Neal_2020, Counterfactual_2024}.
The framework achieves this transformation while ensuring the absence of
confounding. Ultimately, the \emph{fundamental problem of causality} is
reduced to calculating differences between averages of observed outcomes
and the transformed counterfactuals, thereby identifying causal effects.

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
mathematical notation to characterize causal inference, but they do so
in different ways \citep{Neal_2020}. The potential outcomes approach
relies on counterfactual notation, whereas the structural approach
employs do-calculus \citep{Pearl_2009}. Despite these differences, both
notations can be expressed in terms of the other, and both approaches
provide methods for using experimental and observational data to
estimate causal effects \citep{Pearl_2010}.

However, the structural approach offers an additional key advantage over
the potential outcomes approach: it enables the graphical and formal
representation of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020} and and structural causal
models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. DAGs serve as
heuristics, effectively conveying the assumed causal structure of a
system \citep{Pearl_et_al_2016}. A heuristic is a strategy that
simplifies information to make decisions more quickly, efficiently, and
sometimes more accurately than complex methods \citep{Chow_2015}.
Consequently, DAGs do not represent detailed statistical models but
allow researchers to deduce which statistical models can provide valid
causal inferences, assuming the causal structure depicted in the DAGs
are accurate \citep{McElreath_2020}. The Identification-Estimation
flowchart in Figure~\ref{fig-IEflow} visually represents the process of
transitioning from estimands to estimates, and the application of the
sicentific model and data to identify and estimates causal effects.

@@
